---
title: "Correlations DNA vs. markers"
output:
  html_document:
    df_print: paged
---

Here, we investigate the correlations between the expression of the DNA markers and the other markers. 

## Source functions and libraries

```{r}
suppressMessages({
  library(data.table)
  library(ggplot2)
  library(SingleCellExperiment)
  library(CATALYST)
  library(ggrepel)
  library(ggpubr)
  library(RColorBrewer)
  library(ggplotify)
  library(pheatmap)
  library(patchwork)
  source("functions/prep_functions.R")
})

paired_boxes <- function(sce) {
  df <- as.data.table(t(assays(sce)$exprs))
  df[, group := colData(sce)$type]
  df[, patient_id := colData(sce)$patient_id]
  
  df <- melt(df, id.vars = c('group', 'patient_id'), variable.name = 'marker', value.name = 'expression')
  
  df_medians <- df[!group == 'rest', median(expression), by=c("patient_id", "group", "marker")]
  colnames(df_medians) <- c('patient_id', 'group', 'marker', 'expression')
  df_medians[, patient_id := as.factor(patient_id)]
  
  ggpaired(df_medians, x = "group", y = "expression",
           color = "group", line.color = "gray", line.size = 0.4, id="patient_id",
           palette = "jco")+
    #stat_compare_means(paired = TRUE, method = "t.test")+
    facet_wrap(~marker, scales = 'free')
}

dna_boxes <- function(sce) {
  df <- as.data.table(t(assays(sce)$exprs))
  df[, patient_id := colData(sce)$patient_id]
  
  df_melt <- melt(df, id.vars = c('patient_id', 'DNA1', 'DNA2'), variable.name = 'marker', value.name = 'expression')
  DNA_1 <- df_melt[, cor(x=DNA1, y=expression), by=c('patient_id', 'marker')]
  DNA_1[, median_cor := median(V1), by = marker]
  colnames(DNA_1) <- c('patient_id', 'marker', 'correlation', 'median_cor')
  
  DNA_2 <- df_melt[, cor(x=DNA2, y=expression), by=c('patient_id', 'marker')]
  DNA_2[, median_cor := median(V1), by = marker]
  DNA_2 <- DNA_2[order(median_cor)]
  DNA_2[, marker := factor(marker, levels = unique(DNA_2$marker))]
  DNA_1[, marker := factor(marker, levels = unique(DNA_2$marker))]
  colnames(DNA_2) <- c('patient_id', 'marker', 'correlation', 'median_cor')
  
  dna_df <- merge(DNA_1[, -c('median_cor')], DNA_2[, -c('median_cor')], by = c('patient_id', 'marker'))
  colnames(dna_df) <- c('patient_id', 'marker', 'DNA_1', 'DNA_2')
  dna_df <- melt(dna_df, id.vars = c('patient_id', 'marker'), variable.name = 'DNA_marker', value.name = 'correlation')
  
  ggplot(dna_df, aes(x = marker, y = correlation, fill = DNA_marker))+
    geom_boxplot()
}
```

## Global settings
```{r}
theme_set(theme_bw(base_size = 16))
path_to_data <- "/nfs/data/Bongiovanni-KrdIsar-platelets/Cyanus_RPsMPs/data/"
```


## Visualization

Here, we investigate the correlations between expressions of the 2 DNA markers and the other markers for: 

- the unnormalized data
- the data normalized w.r.t. CD42b
- the data normalized w.r.t. CD61
- the data normalized w.r.t. CD41
- the data normalized w.r.t. DNA2

```{r, fig.width = 17, fig.height = 5}
mapping <- c("sce_original_RPs_MPs_rest_DNA.rds", "sce_CD42b_RPs_MPs_rest_DNA.rds", "sce_CD61_RPs_MPs_rest_DNA.rds", "sce_CD41_RPs_MPs_rest_DNA.rds", "sce_DNA2_RPs_MPs_rest_DNA.rds")
names(mapping) <- c('original', 'CD42b', 'CD61', 'CD41', 'DNA2')

for(obj in names(mapping)){
  sce <- readRDS(file.path(path_to_data, mapping[obj]))
  plot(dna_boxes(sce))
  ggsave(paste0('plots/dna_boxes_', obj, '.png'), height=5, width=17)
}
```

Here, we show the pairwise median expressions for RPs vs. MPs for:

- the unnormalized data
- the data normalized w.r.t. CD42b
- the data normalized w.r.t. CD61
- the data normalized w.r.t. CD41
- the data normalized w.r.t. DNA2

```{r, fig.width = 12, fig.height = 8}
mapping <- c("sce_original_RPs_MPs_rest_DNA.rds", "sce_CD42b_RPs_MPs_rest_DNA.rds", "sce_CD61_RPs_MPs_rest_DNA.rds", "sce_CD41_RPs_MPs_rest_DNA.rds", "sce_DNA2_RPs_MPs_rest_DNA.rds")
names(mapping) <- c('original', 'CD42b', 'CD61', 'CD41', 'DNA2')

for(obj in names(mapping)){
  sce <- readRDS(file.path(path_to_data, mapping[obj]))
  plot(paired_boxes(sce))
  ggsave(paste0('plots/paired_boxes_', obj, '.png'), height=8, width=12)
}
```

